TODO: title

### 2014-05-05 paper/main.tex
improved introduction, I think.

  % new user experiance is only chance to make all important first impression
  One of the important problems in recommender systems is the new user experience.
  Every potential user of the system must at some point be a new user.
  If the system cannot provide a good user experience, then the user is likely to leave and not return.
  Even if the user does choose to enter the system, the user's first impression of the system can have a lasting impact on how the user interacts with the system.
  The user's first impression can shape how the user expects the system to behave in future interactions, fundamentally changing how the user perceives the system.

% one important trait is trust.
    % trust is complex
    % matching user's experiance is probably important.
  One important part of how the user interacts with a recommender system is how much the user trusts the recommender system.
  The more the user trusts a recommender system, the more the user can take advantage of it.
  If the user doesn't trust a recommender system, the system can be of no value to the user.
  Building trust with the user is therefore an important part of the new user experience.
  Past work has shown that one of the strongest factors in the formation of trust is how much the system's recommendations match the user's knowledge \cite{wangAttributionOfTrust}.
  For new users this can easily take the form of recommending items that the user has already seen.
  %TODO: I kinda of want to cut this. I want to talk about this theme, but I don't actually know where it fits in.

  % past work tends to focus on the algorithm side, the algorithm side of this problem is cold start.
    % SMAL DATA TENDS TO IMPLY BAD RESULTS.
    % cold start algorithms is a larger problem
    % cold start algorithms tend to use extra information to help imrpove eary performance.
    % we only look at user cold start as it is ubiquitous.
  Past work on new user experience in recommender systems has tended to focus on the algorithmic cold start problem.
  The cold start problem refers to the problem that no algorithm can make useful recommendations with very limited amounts of information.
  While this applies to both new users and new items we will only discuss research on the new user cold start problem.
  Most solutions to the algorithmic cold start problem require some source of extra information about new users.

  % One common new user strategy is to have a special data collection mode.
  % collect ratings fast to get over a precived ``bad period''
  % work on this focuses on how to collect useful ratings.
  % (framework from decision tree work)
  The best studied strategy for new users is to have users go through a new user rating survey.
  Systems that take this practice do not allow their users to receive recommendations until some fixed number of ratings have been entered.
  The idea is that, after enough ratings have been entered, the predictions will be `good enough'.
  This idea has been explored by many authors, most consider a system in which the user is presented with a list of items, and asked to rate items until they have enough ratings.
  For a good discussion of prior work in this direction see \cite{adaptiveBootstrapping}.
  Generally, this work strives to strike a balance between the amount of effort the user must put in before getting enough ratings, and the usefulness of each rating to the user.
  This balance is complicated by how effort effects the user's perception of value \cite{TenIsEnough}.

  % One promising approach is to let user control interaction
  % user or system driven control.
  % promising, we don't know much about follow up.
  One promising approach to the new user rating survey is to have users control which movies they rate.
  This was pioneered by McNee et al., who looked at comparisons between user driven and system driven new user surveys \cite{mcneeInterfaces}.
  They found that users who were in control of which items to rate during a new user survey rated fewer movies, but that the profiles they created were more useful to the system in predicting their future behavior.
  Interestingly, while these users took more time to create their profiles, they reported that the sign up was short at the same rate as the system driven condition.
  While these results are quite promising, to our knowledge no follow up work has been done to explore these issues further.
  This goes to show, however, that the new user experience is a complicated issue and novel strategies may yet prove to be the best solution.

@@

  While past work has covered the new user rating survey as a solution to new user cold start, much less work has been done comparing other strategies.
  Most of this work has been focused on strategies for how to get ratings from users.
  Novel interfaces may elicit information other than ratings such as item comparisons \cite{Loepp}.
  Furthermore, very little work has carefully considered how best to use the information we receive.
  Most authors simply assert that more ratings are better, but for new users we may see interesting performance gains by simply choosing algorithms that perform well for new users.

  We propose to carefully compare the performance of standard algorithms at recommending for new users.
  Our goal is to understand how different algorithms behave for new users, especially how their behavior changes as new users add ratings into the system.
  While there has been much work comparing algorithms in an offline setting, very little of it has explored what are the best algorithms for users with very few ratings.
  Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.

  Our approach is related to temporal evaluations of recommender systems \cite{KorenTemporal, LathiaTemporal, burke2010evaluating}.
  Temporal evaluations tend to evaluate recommender systems as a function of time, seeing how they performed at different points in their lifetime.
  This approach is well suited for understanding the effect of changes to the system, or user base of a recommender.
  Unfortunately, this approach is less well suited towards comparing algorithms, as it necessarily co-mingles the effect of a recommender on user navigation with the quality of the recommender making it hard to consider algorithms that were not deployed with the technique.
  Regardless, we consider this technique an inspiration for our evaluation.
  %TODO: I kind of want to cut this paragraph
  %TODO: Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.

  We structure our work around the following research questions.
  \begin{itemize}
    \item RQ1: How do algorithms behave for users with very few ratings?
    \item RQ2: How does algorithm behavior change with increasing ratings for new users?
  \end{itemize}
  Unfortunately, algorithm behavior is a broad and hard to define subject.
  We will look at three distinct ways to quantify the behavior of a recommender system, each of which may lead to multiple metrics.

@@

  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
  \begin{itemize}
  \item How do algorithms behave with respect to prediction accuracy?
  \item How do algorithms behave with respect to ranking items appropriately?
  \item How do algorithms behave as measured by other metrics such as popularity and diversity?
  \end{itemize}

    % We look at three families of recommender algorithms (user based KNN, item based KNN, SVD)
    % we compare on several dimensions
    % we suggest a methodology for performing these types of computations

@@

>>>>

  % new user experiance is only chance to make all important first impression
  One of the important problems in recommender systems is the new user experience.
  If the system cannot provide a good user experience, then new users are likely to leave and not return.
  Even if the user does choose to enter the system, the user's first impression of the system can have a lasting impact on how the user interacts with the system.
  The user's first impression can shape how the user expects the system to behave in future interactions, fundamentally changing how the user perceives the system.

% one important trait is trust.
    % trust is complex
    % matching user's experiance is probably important.
  %One important part of how the user interacts with a recommender system is how much the user trusts the recommender system.
  %The more the user trusts a recommender system, the more the user can take advantage of it.
  %If the user doesn't trust a recommender system, the system can be of no value to the user.
  %Building trust with the user is therefore an important part of the new user experience.
  %Past work has shown that one of the strongest factors in the formation of trust is how much the system's recommendations match the user's knowledge \cite{wangAttributionOfTrust}.
  %For new users this can easily take the form of recommending items that the user has already seen.
  %TODO: I kinda of want to cut this. I want to talk about this theme, but I don't actually know where it fits in.

  % past work tends to focus on the algorithm side, the algorithm side of this problem is cold start.
    % SMAL DATA TENDS TO IMPLY BAD RESULTS.
    % cold start algorithms is a larger problem
    % cold start algorithms tend to use extra information to help imrpove eary performance.
    % we only look at user cold start as it is ubiquitous.
  Past work on new user experience in recommender systems has tended to focus on the `cold start problem'.
  Cold start refers to the problem that no algorithm can make useful recommendations with very limited amounts of information, therefore new users and items must have worse recommendations.
  We will only discuss the user side of the cold start problem in this work.
  Most solutions to the cold start problem require some source of extra information about new users.

  % One common new user strategy is to have a special data collection mode.
  % collect ratings fast to get over a precived ``bad period''
  % work on this focuses on how to collect useful ratings.
  % (framework from decision tree work)
  The best studied strategy for new users is to have users go through a new user rating survey.
  Systems that take this practice do not allow their users to receive recommendations until some fixed number of ratings have been entered.
  The idea is that after enough ratings have been entered, the predictions will be `good enough'.
  This idea has been explored by many authors, most consider a system in which the user is presented with a list of items, and asked to rate items until they have enough ratings.
  For a good discussion of prior work in this direction see \cite{adaptiveBootstrapping}.
  Generally, this work strives to strike a balance between the effort user needed to get enough ratings, and the usefulness of each rating to the system.
  This balance is complicated by how effort effects the user's perception of value \cite{TenIsEnough}.

  % One promising approach is to let user control interaction
  % user or system driven control.
  % promising, we don't know much about follow up.
  One promising approach to the new user rating survey is to have users control which movies they rate.
  This was pioneered by McNee et al., who looked at comparisons between user driven and system driven new user surveys \cite{mcneeInterfaces}.
  They found that users who were in control of the items they rated made fewer ratings, but that the profiles they created were more useful in predicting the user's future ratings.
  Interestingly, while these users took more time to create their profiles, they were just as likely as other users to report that the sign up was short.
  While these results are quite promising, to our knowledge no follow up work has been done to explore these issues further.
  This goes to show, however, that the new user experience is a complicated issue and novel strategies may yet prove to be the best solution.

@@

  While past work has covered the new user rating survey as a solution to new user cold start, much less work has been done comparing other strategies.
  Most of this work has been focused on strategies for how to get ratings from users.
  Very little work has carefully considered how best to use the information we receive.
  Most authors simply assert that more ratings are better, but for new users we may see interesting performance gains by simply choosing algorithms that perform well for new users.

  We propose to combine two traditions of recommender system research, offline algorithm analysis and new user experience research.
  We seek to compare algorithms based on their performance for new users.
  Our goal is to understand the differences between algorithms for new users, especially how these algorithms respond to different amounts of information.
  While there has been much work comparing algorithms in an offline setting, very little of it has explored what are the best algorithms for users with very few ratings.
  %Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.

  %Our approach is related to temporal evaluations of recommender systems \cite{KorenTemporal, LathiaTemporal, burke2010evaluating}.
  %Temporal evaluations tend to evaluate recommender systems as a function of time, seeing how they performed at different points in their lifetime.
  %This approach is well suited for understanding the effect of changes to the system, or user base of a recommender.
  %Unfortunately, this approach is less well suited towards comparing algorithms, as it necessarily co-mingles the effect of a recommender on user navigation with the quality of the recommender making it hard to consider algorithms that were not deployed with the technique.
  %Regardless, we consider this technique an inspiration for our evaluation.
  %TODO: I kind of want to cut this paragraph
  %TODO: !!! Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.

  We structure our work around the following research questions.
  \begin{itemize}
    \item RQ1: How do algorithms behave for users with very few ratings?
    \item RQ2: How does algorithm behavior change with increasing ratings for new users?
  \end{itemize}
  Unfortunately, algorithm behavior is a broad and hard to define subject.
  We will look at three distinct sub questions to help capture recommender system behavior.
  Each subquestion may represent several actual metrics in our evaluation.
  %We will look at three distinct ways to quantify the behavior of a recommender system. each of which will lead to multiple metrics.

@@

  \begin{itemize}
  \item How do algorithms behave with respect to prediction accuracy?
  \item How do algorithms behave with respect to ranking and recommending?
  \item How do algorithms behave as measured by other metrics such as popularity and diversity?
  \end{itemize}

  We will answer these questions by developing an evaluation framework that can be used to understand how algorithms perform for new users.
  This framework will compare algorithms on eleven different metrics to allow a more complete understanding of the differences between algorithms.
  We will then use this framework to study how three common algorithms behave for new users.
  Finally, we will draw conclusions from our evaluation and outline directions for future work.

@@

  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
